Mean temperature, fluctuation magnitude, and level of biological organization affect biological responses amidst thermal variability: a meta-analysis

### 2021-05-11 writing /thesis_draft/index.Rmd
putting in everyone's edits from orals

---
title: 'Mean temperature, fluctuation magnitude, and level of biological organization affect biological responses amidst thermal variability: a meta-analysis'
author: 'Margaret Anne Slein'
date: 'May 2021'
institution: 'Reed College'
division: 'Mathematics and Natural Sciences'
advisor: 'Sam Fey'
department: 'Biology'
degree: 'Bachelor of Arts'
header-includes:
    - \usepackage{setspace}\onehalfspacing
knit: bookdown::render_book
site: bookdown::bookdown_site
params:
  'Install needed packages for {thesisdown}': True
output:
  thesisdown::thesis_pdf: default 
abstract: 'Ecosystems and organisms have experienced variation on many temporal scales, from diurnal changes in light availability, to seasonal changes in temperature, to decadal changes in weather patterns. Though scientific literature has continued to highlight the importance of ecologically relevant studies for understanding the tenacity and dynamics of ecosystems, there has yet to be a quantitative review of how variability impacts biological responses across taxa and scales of biological organization. Here, we present a quantitative meta-analysis of how thermal variability impacts biological responses across a variety of taxa, organism sizes, and biological organization. Our results suggest that the range of temperatures an organism experiences is the most important driver of response magnitude, with the interaction of mean temperature and fluctuation range emerging equally as significant predictors in our model. Further, our results also suggest that level of biological organization is a less important, though statistically significant, predictor of organismal responses.'
acknowledgements: |

@@

  @alagawany_heat_2017, @delava_effects_2016, @dang_thermal_2019, 
  @singh_effect_2020, @seebacher_embryonic_2014, @whitehead_effect_1989, 
  @kern_temperature_2014, @isaac_isaac_2003pdf_2003, @sunday_thermal_2012, 
  @baojun_seasonal_2014, @fresquet_response_2011, @doyle_survival_1984, 
  @bronikowski_evolutionary_2001, @parachu_marco_new_2017,
  @pina_effect_2003, @simoncini_influence_2019, @dang_thermal_2019, 
  @wang_effects_2015, @du_embryonic_2009, @garcia-ruiz_effects_2011, 
  @glass_should_2019, @kern_physiological_2015-3, @lowenborg_how_2012, 
  @manenti_predictability_2014-2, @pendlebury_variation_2004-1,
  @qu_incubation_2014, @rolandi_costs_2018, @saxon_temperature_2018-1, 
  @semenov_influence_2007, @treidel_temperature_2016-1, @klepsatel_variation_2013,
  @krenek_coping_2012, @tuff_framework_2016

@@

## The historical significance of variability 
It is no secret that life on earth has evolved and persisted in the face of environmental change for many millennia [@bouchard_ecosystem_2014; @bambach_seafood_1993]. Cyanobacteria can trace their origin back nearly 2.5 billion years to a bacteria that evolved the ability to perform oxygenic photosynthesis and changed organismal composition forever [@soo_origins_2017; @blankenship_how_2017]. Cyanobacteria have continued to persist in the world, despite experiencing widely different conditions throughout time [@knoll_chapter_2008]. In contrast, vascular plants have evolved different photosynthetic systems (C4, CAM) to manage a variety of environmental conditions [@guralnick_evolutionary_2007]. Though cyanobacteria and vascular plants evolved new strategies and structures to persist in a new environment, the environment has continued to remain dynamic, such that evolutionary responses and plasticity have become even more relevant [@reed_phenotypic_2010; @burggren_developmental_2018]. Understanding how organisms respond to dynamic environmental conditions is significant for how we recapitulate ecosystems currently and in the future.

## Relevant significance of variability today 
In addition to the importance of environmental variability across geologic timescales, organisms also interact with environmental variability on several shorter timescales. Organisms and the ecosystems they inhabit have experienced variation on many scales, from diurnal changes in light availability, to seasonal changes in temperature, to decadal changes in weather patterns (i.e. El Niño Southern Oscillation).  As such, the scientific literature has continued to highlight the importance of ecologically relevant studies for drawing robust conclusions about the tenacity and dynamics of ecosystems [@khelifa_usefulness_2019; @stewart_mesocosm_2013]. How variability impacts populations may impact biodiversity, such that only certain species may survive, and how that biodiversity is maintained are key components to the full picture of ecosystem dynamics. As environmental patterns continue to change, from increased mean temperatures to more severe weather events (i.e. hurricanes), understanding the limits to organisms’ ability to cope with changes in environmental conditions is increasingly important for both individual level responses as well as how individual level impacts scale across population and community level responses [@helmuth_organismal_2010; @bernardo_biologically_2014; @huey_predicting_2012].

## Why this study on variability matters 
There is a large body of literature on environmental variability including its relevance for positive effects on populations and communities, i.e. inflationary effects in sink-source systems [@holt_impacts_2003; @vasseur_phase-locking_2009], such that variability aids population density when organisms emigrate. However, there is a lack of consensus as to which scales of biological organization, i.e. individuals, populations, or communities of organisms will be most severely affected by different shifts in environmental conditions (problem statement). We aim to better understand how environmental variability affects the magnitude of responses across organization levels as well as important factors implicated in mediating this response (aims). To accomplish this, we review recent progress in the field of environmental variability, specifically changes in temperature, to identify gaps between theory, experiment, and implementation.

>>>>

---
title: 'Mean temperature, fluctuation magnitude, and level of biological organization affect biological responses amidst thermal variability: a meta-analysis'
author: 'Margaret Anne Slein'
date: 'May 2021'
institution: 'Reed College'
division: 'Mathematical and Natural Sciences'
advisor: 'Sam Fey'
department: 'Biology'
degree: 'Bachelor of Arts'
header-includes:
    - \usepackage{setspace}\onehalfspacing
knit: bookdown::render_book
site: bookdown::bookdown_site
params:
  'Install needed packages for {thesisdown}': True
output:
  thesisdown::thesis_pdf: default 
abstract: 'Ecosystems and organisms have experienced variation on many temporal scales, including diurnal changes in light availability, seasonal changes in temperature, and decadal changes in weather patterns. Though scientific literature has continued to highlight the importance of ecologically relevant studies for understanding the persistence and dynamics of ecosystems, there has yet to be a quantitative review of how variability impacts biological responses across taxa and scales of biological organization. Here, we present a quantitative meta-analysis of how thermal variability impacts biological responses across multiple taxa, organism sizes, and levels of biological organization. Our results suggest that the range of temperatures an organism experiences is the most important driver of response magnitude, with the interaction of mean temperature and fluctuation range emerging equally as significant predictors in our model of biological responsiveness to temperature variability. Further, our results also suggest that level of biological organization is a less important, though statistically significant, predictor of organismal responses.'
acknowledgements: |

@@

  @alagawany_heat_2017, @delava_effects_2016, @dang_thermal_2019, 
  @singh_effect_2020, @seebacher_embryonic_2014, @whitehead_effect_1989, 
  @kern_temperature_2014, @isaac_isaac_2003pdf_2003, @sunday_thermal_2012, 
  @baojun_seasonal_2014, @fresquet_response_2011, @doyle_survival_1984, 
  @bronikowski_evolutionary_2001, @parachu_marco_new_2017,
  @pina_effect_2003, @simoncini_influence_2019, @dang_thermal_2019, 
  @wang_effects_2015, @du_embryonic_2009, @garcia-ruiz_effects_2011, 
  @glass_should_2019, @kern_physiological_2015-3, @lowenborg_how_2012, 
  @manenti_predictability_2014-2, @pendlebury_variation_2004-1,
  @qu_incubation_2014, @rolandi_costs_2018, @saxon_temperature_2018-1, 
  @semenov_influence_2007, @treidel_temperature_2016-1, @klepsatel_variation_2013,
  @krenek_coping_2012, @tuff_framework_2016, @vasseur_increased_2014, @khelifa_usefulness_2019

@@

## The historical significance of variability 
It is no secret that life on earth has evolved and persisted in the face of environmental change for many millennia [@bouchard_ecosystem_2014; @bambach_seafood_1993]. Cyanobacteria can trace their origin back nearly 2.5 billion years to a bacteria that evolved the ability to perform oxygenic photosynthesis and changed organismal composition forever [@soo_origins_2017; @blankenship_how_2017]. Cyanobacteria have continued to persist in the world, despite experiencing widely different conditions throughout time [@knoll_chapter_2008]. Vascular plants have also evolved different photosynthetic systems (e.g. C4, CAM) to manage a variety of environmental conditions [@guralnick_evolutionary_2007]. Though cyanobacteria and vascular plants evolved new strategies and structures to persist as the environment changed, the environment has continued to remain dynamic, such that evolutionary responses and plasticity have become even more relevant [@reed_phenotypic_2010; @burggren_developmental_2018]. Understanding how organisms respond to dynamic environmental conditions is significant for how we recapitulate ecosystems currently and in the future.

## Relevant significance of variability today 
In addition to the importance of environmental variability across geologic timescales, organisms also interact with environmental variability on several shorter timescales. Organisms and the ecosystems they inhabit have experienced variation on many scales, including diurnal changes in light availability, seasonal changes in temperature, and decadal changes in weather patterns (e.g. El Niño Southern Oscillation).  As such, the scientific literature has continued to highlight the importance of ecologically relevant studies for drawing robust conclusions about the dynamics of ecosystems [@khelifa_usefulness_2019; @stewart_mesocosm_2013]. How variability impacts populations may impact biodiversity, such that only certain species may survive, and how that biodiversity is maintained are key components to the full picture of ecosystem dynamics. As environmental patterns continue to change, from increased mean temperatures to more severe weather events (e.g. hurricanes), understanding the limits to organisms’ ability to cope with changes in environmental conditions is increasingly important for both individual level responses as well as how individual level impacts scale across population and community level responses [@helmuth_organismal_2010; @bernardo_biologically_2014; @huey_predicting_2012].

## Why this study on variability matters 
There is a large body of literature on environmental variability including its relevance for positive effects on populations and communities, e.g. inflationary effects in sink-source systems [@holt_impacts_2003; @vasseur_phase-locking_2009], such that variability aids population density when organisms emigrate. However, there is a lack of consensus as to which scales of biological organization, e.g. individuals, populations, or communities of organisms, will be most severely affected by different shifts in environmental conditions. We aim to better understand how environmental variability affects the magnitude of responses across organization levels as well as important factors implicated in mediating this response. To accomplish this, we review recent progress in the field of environmental variability, specifically changes in temperature, to identify gaps between theory, experiment, and implementation.
